Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019

### 2020-11-05 manuscript/EP1162_manuscript_draft.Rmd
init

@@

 # Create figure showing marginal means from model
fitBA.Ann.plot<-
  ggpredict(fitBA.Ann,terms=c("Year","Species","BA.Target"))  %>%
  plot()+
  theme_bw()+
  ylab("Predicated annualized basal area increment (m²)")+
  xlab("Measurement year")+
  ggtitle("")+
  theme(legend.position = "bottom")

>>>>

  # piDat.table %>% 
  #   ggplot(aes(x=diamClass,y=BAImean,color=BA.Target))+
  #   geom_point()+
  #   facet_wrap(~Species)

@@

 # Create figure showing marginal means from model
fitBA.Ann.plot<-
  ggpredict(fitBA.Ann,terms=c("Year","Species","BA.Target"))  %>%
  plot()+
  theme_bw()+
  ylab("Predicted annualized basal area increment (m²)")+
  xlab("Measurement year")+
  ggtitle("")+
  theme(legend.position = "bottom")

### 2020-11-05 manuscript/EP1162_manuscript_draft_October2020.Rmd
init

# Results

## Tree-level

### Basal area increment from 1992-2019

@@

@@

----------------

# Create dead dataset

# Dead species totals
deadList<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  group_by(ID) %>% 
  summarise_all(first)

dead<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
  group_by(ID) %>% 
  summarise_all(first) %>% 
  ungroup() %>% 
  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
  group_by(BA.Target,DBH.class,Year) %>% 
  summarise(BA=sum(BA),
            k=n())

dead.Summ<-
  dead %>% 
  ungroup() %>% 
  group_by(BA.Target) %>% 
  summarise(BA=sum(BA)/sum(dead$BA)*100,
            k=sum(k)/sum(dead$k)*100) %>% 
  mutate_if(is.numeric,round,1)

p1<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=k,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total tree mortality (stems)")+
  facet_wrap(~BA.Target)+
  xlab("")+
  theme(axis.text.x = element_blank())

p2<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=BA,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total basal area mortality (m²)")+
  facet_wrap(~BA.Target)

mort.plot<-cowplot::plot_grid(p1, p2, ncol = 1, align = "v")
figNums("mort.plot",display=FALSE)

A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died from 1992 to 2019.  No tree mortality was noted during the 1992 and 1994 data collection.  By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).  Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
`r figNums("mort.plot",display="cite")`).

```{r display mortality plot}
mort.plot
```

`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1992-2019, presented as total stems (top panel) and basal area (bottom panel).")`

From Mike's email:
Eventually as BA stand density increases, density dependent mortality kicks in, because there is only so much growing space or access to site resources to go around. This can occur with the stand (conceptually or hypothetically)  in two possible ways: (a) “classic” suppression of smaller tree classes through asymmetric competition of bigger trees shading out smaller ones, or (b) high stand densities suppressing the vigor of the whole stand through various means, suggesting more symmetric competition – i.e. the idea that small trees compete for growing space with big ones as well. Asymmetric competition is generally accepted by most foresters and silviculturists; symmetric competition less so, and is often harder to demonstrate.

*For the discussion: it seems that we're seeing mortality increasing in the plots, especially the control unit over the last 11 years.  Does this reflect the density-dependent mortality that Mike refers to above?.*

More from Mike on mortality:

The thing about unevenaged or complex stand management, esp in ‘wilder’ BC stands, is that we are working in stands with a long legacy of pre-existing ecosystem processes before we enter the stand to manipulate them. Clearcutting liquidates these legacies for the most part, and traditional silviculture and GY has not had to deal with mortality much, other than clearly density dependent suppression of smaller dbh classes. [Side issue: most BC experience and data with unevenaged stands has been in IDF, which have been fire excluded on dry sites, and unevenaged structures allowed to develop. We must be careful about comparing Summit data to IDF precedents, because of substantial differences in ecology.]

And we are working a more extensive level of management in the SBS and ESSF,  where our stands are older, and not manicured nor vacuumed up for mortality and blowdown. From an ecosystem management point of view, that Is probably a good thing.

So warts and all, I think mortality is an inherent characteristic of the types of spruce-balsam stands and management we are studying. Expecting and documenting mortality was built into the EP 1162 working plan. The treatment effects we anticipate are the net stand growth = gross growth minus mortality.

Conceptually, tree mortality following the 1991/92 harvest (and subsequent sanitation felling)  functions as a BA removal from the live-tree pool. In most plots, BA of trees that die within a measurement period is subsumed or overridden by the total BA growth of the rest of the trees during that period.

Where the periodic data trends get messier is where single big trees or a cluster of trees gets killed by something, resulting in a pulse of BA mortality. Examples: The great big Fdi in Plot 18, or notably, about 3-4 big Sx in Plot 3 that got nailed by spruce beetle.

In our results and discussion:
1.	We can report on what stand growth and development trends seem to be clearly and unambiguously  influenced by initial BA conditions, and the magnitude of these treatment effects. 
2.	A component of tree mortality (e.g. certain size classes) may be density dependent. 
3.	A component of tree mortality may be density-independent (a.k.a stochastic or chaotic). 
4.	By reporting on all the above, we are providing managers with results that integrate silvicultural expectations observations from real-life “messy” stands which will experience all of the above. By keeping our eye on overall stand growth and performance following treatment, we can inform / reassure managers that individual tree mortality events are not the end of the world, but rather need to be viewed in the context of overall stand dynamics and growth expectations.

## Stand-level analyses

@@

appNums("htModelFit.plot",display="FALSE") # model output

@@

### Estimating tree height from diameter
Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (i.e., p>0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effects. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log-transformed DBH (not shown).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit.plot",display="cite")`). This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].

----------------------------
```{r height model fit}

htModelFit.plot

`r appNums("htModelFit.plot",caption="Scatterplot of predicted height(m) vs actual height(m) from diameter-height model.")`

@@

@@

# diamClass.plot<-cowplot::plot_grid(diamVol.plot,diamBA.plot,diamK.plot,ncol=1)

@@

### Plot means

```{r generate stand level plot means}

plotData.X<-
# Determine which plots have measurements in 1992 and 2019
  X %>% 
  #  filter(Year %in% c(1992,2019)) %>% 
  #  filter(Status=="Live") %>% 
  # 
  # group_by(Plot,Year) %>% 
  # summarise(n=n()) %>% 
  # ungroup() %>% 
  # group_by(Plot) %>% 
  # summarise(n=length(Year)) %>% 
  # filter(n>1) %>% # pick plots with measurements in both years
  # pull(Plot) %>% 
  # as.data.frame() %>% 
  # setNames("Plot") %>% 
  # 
  # # Now filter original dataset 
  # left_join(X,by="Plot") %>% 
  # mutate(Plot=fct_drop(Plot)) %>%  # drop unused levels
  # 
  # # Filter for years of interest
  # filter(Year %in% c(1992,2019)) %>% 
  # mutate(Year=fct_drop(Year)) %>% 
  filter(Status=="Live") %>%

  # Remove volumes below 18.5cm DBH
  mutate(vol=replace(vol,DBH<18.5,0)) %>%

  # now create plot sums grid so that you can calculate means
  group_by(BA.Target,Plot,Year) %>% 
    summarise(Vol.sum=sum(vol,na.rm=T)*20,
              BA.sum=sum(BA)*20,
              SPH.sum=n()*20) %>%

  ungroup() %>%

  # Create QMD
  mutate(QMD.mean=sqrt(BA.sum/(SPH.sum*0.0000785)))

# Now summarise plot data at the treatment unit level
standData.X<-
  plotData.X %>% 
  group_by(BA.Target,Year) %>% 
  summarise(Vol=mean(Vol.sum),
            Vol.se=sd(Vol.sum)/sqrt(length(Vol.sum)),
            BA=mean(BA.sum),
            BA.se=sd(BA.sum)/sqrt(length(BA.sum)),
            SPH=mean(SPH.sum),
            SPH.se=sd(SPH.sum)/sqrt(length(SPH.sum)),
            QMD=mean(QMD.mean),
            QMD.se=sd(QMD.mean)/sqrt(length(QMD.mean)))

# PLotting

# Create function
standPlot<-function(standVar,y.label) {

    standData.X %>% 
    mutate(Year=as.numeric(as.character(Year))) %>% 
    dplyr::select(BA.Target,Year,standVar,paste0(standVar,".se")) %>% 
    setNames(c("BA.Target","Year","Mean","SE")) %>%

    ggplot(aes(x=Year,y=Mean,color=BA.Target,group=BA.Target))+
    geom_point()+
    geom_line()+
    geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
                   position=position_dodge(.9))+
    facet_wrap(~BA.Target)+
      scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "none")+
    ylab(y.label)

}

standVol.plot<-
  standPlot("Vol",y.label="Vol. (m³/ha)")+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())

standBA.plot<-
  standPlot("BA",y.label="BA (m²/ha)")+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())

standK.plot<-
  standPlot("SPH",y.label="Tree dens. (sph)")+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())

standQMD.plot<-standPlot("QMD",y.label="QMD (cm)")

standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,standQMD.plot,ncol=1,
                      align="v")

figNums("standPlots",display=FALSE)
```

```{r fit model to plot data}

# Rename columns in plotData.X
plotData.X<-
  plotData.X %>% 
  rename(Vol="Vol.sum",
         BA="BA.sum",
         SPH="SPH.sum",
         QMD="QMD.mean")

# fit linear models
fit.vol<-lmer(Vol~BA.Target*Year+(1|Plot),data=plotData.X)
fit.ba<-lmer(BA~BA.Target*Year+(1|Plot),data=plotData.X)
fit.sph<-lmer(SPH~BA.Target*Year+(1|Plot),data=plotData.X)

# Summarize models
library(lmerTest)
anova(fit.vol, type="3") 
  rbind(anova(fit.ba, type="3"))
anova(fit.sph, type="3")

# Create estimated marginal means
ba.means <- emmeans(fit.ba, ~ BA.Target*Year)
sph.means <- emmeans(fit.sph, ~ BA.Target*Year)

# Contrasts
ba.contrast <- contrast(ba.means,interaction=c("pairwise","revpairwise")) 
sph.contrast <- contrast(sph.means,interaction=c("pairwise","revpairwise"))
```

Plot means of volume and basal area show that from 2009-2019, stand volume and basal area converged between both harvested treatment units, and were less than the control (
`r figNums("standPlots",display="cite")`).  The slope of all attributes showed that the low RBA accrued volume, density and diameter faster than the other units over the 27-year period of this study.  A linear model

-----------

```{r fig.height=10}

standPlots
```

`r figNums("standPlots",caption="Mean plot attributes by treatment unit and year.  Whiskers are one standard error of the mean.  Vol., volume; BA, basal area; Tree dens., tree density; QMD, quadratic mean diameter")`

*Note that sample size varies with year.  Some plots were not measured in 1992.  This likely explains the dip in basal area and volume in TU:20 from 1992 to 1994, despite no recorded mortality in 1992/1994.*

### Stand attribute by diameter class

Among treatments and throughout the stand development after treatment, the stands have maintained a reverse-J diameter class distribution (`r figNums("diamK.plot",display="cite")`).  
Immediately after the partial harvest, the diameter class distribution of volume differed between the treatment.  In the 27 years since partial harvest, volume in all three treatment units was mostly concentrated in the 22.5-27.5cm DBH class, and mostly comprised of fir (`r figNums("diamVol.plot",display="cite")`).  Over both time periods shown in the figure and all three treatments, most of the spruce volume was concentrated in DBH classes spanning 37.5 to 47.5cm.  Basal area trends were very similar to volume, and are not shown.

-----------

```{r}
diamK.plot

`r figNums("diamVol.plot",caption="Diameter-class distribution of tree density (stems/hectare) by species, treatment and year.")`

```{r}
diamVol.plot

`r figNums("diamVol.plot",caption="Diameter-class distribution of volume (m³/hectare) by species, treatment and year.")`

*Note that in this figure, we show volume for all diameter classes.  This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*

### Species composition

```{r summarize changes in species composition}

@@

figNums("sppVol.plot",display=FALSE)

@@

A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce were organized similarly between the high RBA and control units (`r figNums("sppVol.plot",display="cite")`).  In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.  This pattern was found for volume, basal area, and tree density.  We only show volume in this document.  By 2019, the fir-spruce composition by distribution class was maintained in these two units, with small increases in spruce composition in the smallest diameter class.

In the low RBA, the species composition distribution was different.  Here, both species maintained a similar proportion of fir-leading across diameter classes smaller than 37.5cm DBH, whereas spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.

```{r species composition plot}
sppVol.plot

@@

`r figNums("sppVol.plot",caption="Spruce-fir volume proportional composition by diameter class, treatment and year.")`

*Note this figure is almost identical if we compare stems, basal area or volume. For brevity, we only show volume here.*

>>>>

---
title: Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019 
author: "Mike Jull and Hardy Griesbauer"
date: "29/03/2020"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    collapsed: false
    theme: united
    number_sections: true
  word_document:
    reference_docx: wordTemplate2.docx
csl: cjfr.csl
bibliography: ep1162Bib.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)

# Load libraries
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)
library(cowplot)
library(officer)
library(sjPlot)
library(emmeans)

# Load data
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

# SOME PRELIM DATA FORMATTING
# formatting plotDat
plotDat<-
  plotDat %>% 
  mutate(Plot=as.factor(Plot))

# set up diameter classes
diamClass=seq(from=7.5,to=100,by=5)

# Set against scientific notation
options(scipen=999)

# Merge data
X<-
  dat %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(DBH.class=cut(DBH,diamClass,include.lowest=TRUE,ordered_result = TRUE,right=F,
                       labels=diamClass[1:(length(diamClass)-1)])) %>% 
  mutate(Plot=as.factor(Plot)) %>%

  # replace species codes with words
  mutate(Species=recode_factor(Species,Bl="Fir",Sx="Spruce")) %>% 
  mutate(Species=fct_drop(Species)) %>% # have an extra level in here for some reason

  # Create TU lookup
  mutate(BA.Target=paste0("TU:",BA.Target))

# set up captions

# Code to create captions for tables
tableNums <- captioner(prefix = "Table")

# Code to create captions for figures
figNums <- captioner(prefix="Figure")

# Code to create captions for appendices
appNums <- captioner(prefix="Appendix")

# Set ggplot theme
theme_set(theme_bw())

# Define function to convert diameter (cm) to basal area (m2), and vice versa
d2ba<-function(diam) {return((diam/2/100)^2*pi)}
ba2d<-function(ba) {return(sqrt(ba/pi)*200)}

# Methods

```{r define captions to be used in methods results and discussion}

# Captions (put these in order)

  # Introductory 
  appNums(name="paiTable",caption="Number of growing seasons between each measurement period",display=FALSE) # Table showing number of growing seasons between each measurement period
  appNums(name="standVol.plot",caption="Plot total volume for five measurement periods",display=FALSE) # Plot volume figure
  appNums(name="standBA.plot",caption="Plot basal area for five measurement periods",display=FALSE)  # Plot basal area figure
  appNums(name="standQMD.plot",caption="Plot quadratic mean diameter for five measurement periods",display=FALSE) # QMD figure  
  appNums(name="standK.plot",caption="Plot tree density for five measurement periods", display=FALSE) # SPH figure

  # Tree BAI model
  figNums(name="treeBAI.plot",display=FALSE)
  tableNums("fitBA.table",display=FALSE) # Summary of mixed effects model for BAI
  appNums("sppContrast.table",display=FALSE) # Species contrasts from treeBAI model
  appNums("trtContrast.table",display=FALSE) # Treatment contracts from treeBAI model

  # Tree heights model
  appNums("heightDiamRawPlot",display=FALSE) # Plot of diameter/heights`
  appNums("htModelFit",display="FALSE") # model output

  # Diameter class analysis
  figNums("diamVol.plot", caption="Mean plot volume distribution by diameter class and treatment, for two periods.  Black line shows total volume.  Volume calculated on trees meeting 18.5cm DBH limit.  Spruce and fir contribution to volume by diameter class are shown with bar plots.  Douglas-fir volume contribution by diameter class is not shown.",display="FALSE")

  figNums("diamBA.plot", caption="Mean plot basal area distribution by diameter class and treatment, for two periods.  Black line shows total basal area.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

  figNums("diamK.plot", caption="Mean tree density distribution by diameter class and treatment, for two periods.  Black line shows total tree density.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

## Analyses

### Tree-level growth responses
To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects, and plot as a random intercept effect.  We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].

### Stand structure and growth dynamics

#### Stand-level volume and basal area estimates
Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.  Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.  For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.  Trees with broken tops were omitted from diameter-height models. We summed tree-level estimates of volume and basal area to generate stand-level estimates.

#### Stand growth

To test the hypothesis that stand volume and basal area growth from 1992-2019 varied by treatment, we fit separate linear mixed effects models using periodic annual volume and basal area increments as response variables, treatment as a fixed effects, and plot as a random effect.  Periodic annual increment was calculated for the four measurement periods by dividing increment over the measurement period by the number of complete growing seasons within that period (`r appNums("paiTable",display="cite")`).

## Results

@@

### Tree-level basal area increment from 1992-2019

@@

## Stand-level development over 27 years

@@

@@

### Estimating tree heights
Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (i.e., p>0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit",display="cite")`). This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].

@@

```{r generate stand attributes}

standDatspp<-

# Create tree volumes
  X %>% 
  filter(Status=="Live") %>% # filter for live trees
  filter(DBH>=18.5) %>% # filter for trees with minimum diameter utilization limits
  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  dplyr::select(Plot,TreeID,Year,vol) %>%

# Create stand data, by species
  right_join(X,by=c("Plot","TreeID","Year")) %>%
  group_by(BA.Target,Plot,Species,Year) %>% 
  mutate(BA=d2ba(DBH)) %>% # create BA column
  summarise(Vol=sum(vol,na.rm=T)*20, # notice that all variables are by hectare
            BA=sum(BA,na.rm=T)*20,
            k=n()*20) %>% 
   mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
  arrange(BA.Target,Plot,Species,Year) %>% 
  ungroup()

# Create stand data with totals
standDat<-
# Create tree volumes
  X %>% 
  filter(Status=="Live") %>% # filter for live trees
  filter(DBH>=18.5) %>% # filter for trees with minimum diameter utilization limits
  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  dplyr::select(Plot,TreeID,Year,vol) %>%

# Create stand data, by species
  right_join(X,by=c("Plot","TreeID","Year")) %>%
  group_by(BA.Target,Plot,Year) %>% 
  mutate(BA=d2ba(DBH)) %>% # create BA column
  summarise(Vol=sum(vol,na.rm=T)*20, # notice that all variables are by hectare
            BA=sum(BA,na.rm=T)*20,
            k=n()*20) %>% 
   mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
  mutate(Species="Total") %>% 
  dplyr::select(BA.Target,Plot,Species,Year,Vol,BA,k,QMD) %>% 
  ungroup() %>% 
  rbind(standDatspp) %>% 
  arrange(BA.Target,Plot,Species,Year) %>% 
  filter(Species %in% c("Total","Spruce","Fir")) # limit to certain species

####
# Basic plotting function
standDat.plot<-function(standVar,y.label) {
  standDat %>% 
  mutate(Plot=paste0("Plot:",Plot)) %>% 
  mutate(Year=as.numeric(as.character(Year))) %>% 
  ggplot(aes(x=Year,y=get(standVar),color=Species,group=Species))+
  geom_point()+
  geom_line()+
    ylab(y.label) +
   facet_wrap(BA.Target~Plot,ncol=5)+
   scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "bottom")+
   scale_color_manual(values=c("#E69F00", "#56B4E9","#000000")) %>%  # control color

  return()
} # close function

# create stand-level plots
standVol.plot<-standDat.plot("Vol",y.label="Total volume (m³/ha)")
standBA.plot<-standDat.plot("BA",y.label="Basal area (m²/ha)")  
standQMD.plot<-standDat.plot("QMD",y.label="Quadratic mean diameter (cm)")  
standK.plot<-standDat.plot("k",y.label="Total tree density (sph)")

### Stand-level attributes by diameter class
To elucidate how stand structure varied among treatments, we also evaluated stand attributes across diameter classes.  Diameter classes were produced in 5cm increments.

@@

@@

### Species composition
```{r summarize changes in species composition}

@@

@@

# Create dead dataset
dead<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  group_by(ID) %>% 
  summarise_all(first) %>% 
  ungroup() %>% 
  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
  group_by(BA.Target,DBH.class) %>% 
  summarise(BA=sum(BA),
            k=n())

p1<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=k))+
  geom_bar(stat="identity")+
  ylab("Total tree mortality (stems)")+
  facet_wrap(~BA.Target)+
  xlab("")+
  theme(axis.text.x = element_blank())

p2<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=BA))+
  geom_bar(stat="identity")+
  ylab("Total basal area mortality (m²)")+
  facet_wrap(~BA.Target)

cowplot::plot_grid(p1, p2, ncol = 1, align = "v")

@@

## TABLES

## FIGURES

```{r plot volume by diameter class}

diamVol.plot
```

`r figNums("diamVol.plot")`

----------------------

```{r plot basal area by diameter class}

diamBA.plot

`r figNums("diamBA.plot")`

```{r plot tree density by diameter class}
diamK.plot
```

`r figNums("diamK.plot")`

---------------------

## APPENDICES/SUPPLEMENTARY MATERIALS

`r appNums("paiTable",caption="Number of growing seasons between measurement periods")`
```{r show paiTable, ft.align="left"}

  tibble(`Measurement Year`=c(1992,1994,1997,2009,2019),`Growing Seasons`=c(NA,3,3,11,11)) %>% 
  flextable() %>% 
  align(align="left",part="all") %>% 
  # set_caption(caption="Number of growing seasons between measurement periods") %>% 
  autofit()

------------------------

```{r set figure options for Appendix figs}
# set options for figures in the next few sections
knitr::opts_chunk$set(fig.width=8, fig.height=8)
```

```{r plot volume}

standVol.plot
```

`r appNums("standVol.plot")`

----------------------

```{r plot basal area}
standBA.plot
```

`r appNums("standBA.plot")`

```{r qmd plot}
standQMD.plot
```

`r appNums("standQMD.plot")`

-----------------

```{r SPH plot}
standK.plot
```

`r appNums("standK.plot")`

------------------------

## REFERENCES
